---
title: "Mapping Antimicrobial Resistance Across U.S. States"
subtitle: "An Introduction to GIS Methods for Environmental Pharmacoepidemiology"
author: "ISPE 2026 Pre-Conference Course"
format:
  revealjs:
    theme: simple
    slide-number: true
    scrollable: true
    code-line-numbers: false
    fig-width: 7.5
    fig-height: 4.2
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Welcome

## Datasets

Open source CDC surveillance data.

No shapefile, no lat/long provided.

- `ARPSP_...csv` — MDR *E. coli* resistance among HAIs, by state, 2024 → **clinical outcome**
- `BEAM_Dashboard_...csv` — enteric pathogen isolates, by state, month, and source (animal/food/human/**environmental**), 2018– → **exposure / surveillance-effort**
- `2024-SIR-ACH.xlsx` — NHSN hospital SIR report, 2024 → used in the independent-practice section

::: notes
Emphasize: most real-world exposure data in this field arrives as tables tied to an area + a time period, not points on a map.
:::

## Learning objectives {.smaller}

- Import and clean messy public-health surveillance data
- Get U.S. state polygons; understand what a CRS is doing for you
- The **attribute join** — non-spatial table → spatial geometry
- Choropleth mapping with `ggplot2` + `sf`
- **Small-multiples panels** across years and seasons — and why the color scale must be shared
- Reading **seasonality** in area-level data
- Understanding how area-level analysis misclassification may impact your results : **MAUP**, **ecological fallacy**, **surveillance-effort bias**

------------------------------------------------------------------------

# Setup

## Packages {.smaller}

```{r load-packages}
library(sf); library(tigris); library(dplyr); library(tidyr)
library(readr); library(readxl); library(janitor)
library(ggplot2); library(viridis); library(scales)

options(tigris_use_cache = TRUE)
sf::sf_use_s2(TRUE)
```

> `tigris` downloads state boundaries once, then caches.
>
> No internet?
>
> Use the offline `cb_2025_us_state_20m` shapefile instead (see instructor note).

------------------------------------------------------------------------

# Part 1 — Import & clean

## 1a. Clinical outcome: MDR *E. coli* {.smaller}

```{r import-arpsp}
ecoli <- read_csv("../data/ARPSP_MDR-Ecoli---Percent-Resistant-in-2024.csv") %>%
  clean_names() %>%
  select(state, percent_resistant) %>%
  mutate(percent_resistant = as.numeric(percent_resistant))

glimpse(ecoli)
```

::: notes
**Ask the room:** how many rows? One per state? Where are PR / DC?
:::

## 1b. Environmental signal: BEAM {.smaller}

Semicolon-delimited; `Year` has thousands-separator commas — typical of real exports.

```{r import-beam}
beam <- read_delim("../data/BEAM_Dashboard_-_Report_Data_20260804.csv",
                    delim = ";", show_col_types = FALSE) %>%
  clean_names() %>%
  mutate(year = as.numeric(gsub(",", "", year)))

valid_state_codes <- c(state.abb, "DC")

beam_env_clean <- beam %>%
  filter(year == 2024, source_type == "Environment", pathogen == "Salmonella",
         state %in% valid_state_codes) %>%
  group_by(state) %>%
  summarize(env_isolates = sum(number_of_isolates, na.rm = TRUE), .groups = "drop")
```

## Discussion: 1b {.smaller}

Only \~20 states show up here, for 2024 alone.

> Is that because *Salmonella* isn't in the environment elsewhere — or because those states aren't sampling it?

An "exposure" layer built from surveillance data always measures exposure **and** *where people looked*.

------------------------------------------------------------------------

# Part 2 — Get the map

## State boundary polygons

```{r get-states}
states_sf <- tigris::states(cb = TRUE, resolution = "20m", year = 2022) %>%
  clean_names() %>%
  select(name, stusps, geoid, geometry)
```

```{r plot-blank-map, fig.height=3.6}
ggplot(states_sf) + geom_sf() + theme_minimal() +
  labs(title = "U.S. state boundaries (no data joined yet)")
```

## Why CONUS-only? {.smaller}

AK / HI / PR sit far outside the CONUS — they'd get cut off or shrink everything else. (`tigris::shift_geometry()` can inset them later — out of scope today.)

```{r conus-only}
conus_codes <- setdiff(c(state.abb, "DC"), c("AK", "HI"))

states_conus <- states_sf %>%
  filter(stusps %in% conus_codes) %>%
  st_transform(5070)   # Albers Equal Area — projected CRS for CONUS
```

**Coordinate Reference System** (**CRS) check-in:** `tigris` gives unprojected NAD83. EPSG:5070 is equal-area, built for CONUS. Always ask: *what CRS is this, and does it fit what I'm about to compute?*

------------------------------------------------------------------------

# Part 3 — Attribute join

## Attaching tables to geometry {.smaller}

Shared key: full state name (ARPSP) / two-letter code (BEAM).

```{r attribute-join}
map_data <- states_conus %>%
  left_join(ecoli, by = c("name" = "state")) %>%
  left_join(beam_env_clean, by = c("stusps" = "state")) %>%
  mutate(env_isolates = coalesce(env_isolates, 0))
```

**Always check your join** before trusting a map:

```{r check-join}
map_data %>% st_drop_geometry() %>%
  filter(is.na(percent_resistant)) %>% select(name, stusps)
```

A silent join failure (trailing space, "St." vs "Saint") is the #1 cause of a wrong-looking choropleth.

------------------------------------------------------------------------

# Part 4 — Your first choropleth maps

## Map: clinical resistance

```{r choropleth-ecoli, fig.height=4}
ggplot(map_data) +
  geom_sf(aes(fill = percent_resistant), color = "white", linewidth = 0.15) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "% MDR\nE. coli") +
  theme_void() +
  labs(title = "Multidrug-resistant E. coli among HAIs, by state, 2024",
       subtitle = "GREAT: continuous, red = higher resistance",
       caption = "Source: ARPSP")
```

## Map: environmental signal

```{r choropleth-beam, fig.height=4}
ggplot(map_data) +
  geom_sf(aes(fill = env_isolates), color = "white", linewidth = 0.15) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "% MDR\nSalmonella") +
  theme_void() +
  labs(title = "Environmentally-sourced Salmonella isolates, by state, 2024",
       subtitle = "CDC BEAM enteric bacteria surveillance",
       caption = "Source: BEAM Dashboard")

```

<!--# Does this look like contamination — or like where people tested? -->

## On intuitiveness ... palette choice? {.smaller}

Lets plot the same variable (`percent_resistant`), using three different color decisions.

## A qualitative palette on ordered data

```{r map-bad, fig.height=5.8}
map_data_bad <- map_data %>%
  mutate(resist_bin = cut(percent_resistant, breaks = c(0,5,10,15,25),
                           labels = c("0-4%","5-9%","10-14%","15%+"), include.lowest = TRUE))

ggplot(map_data_bad) +
  geom_sf(aes(fill = resist_bin), color = "white", linewidth = 0.15) +
  scale_fill_brewer(palette = "Set1", name = "% MDR\nE. coli") +
  theme_void() + labs(subtitle = "Colors imply no order")
```

<!--# Red isn't obviously "more" than blue or green — reader has to keep re-checking the legend. -->

## Not so great: right idea, wrong bins

```{r map-meh, fig.height=5.8}
map_data_meh <- map_data %>%
  mutate(resist_bin = cut(percent_resistant, breaks = c(0,20,40,60,80,100),
                           labels = c("0-19%","20-39%","40-59%","60-79%","80-100%"),
                           include.lowest = TRUE))

ggplot(map_data_meh) +
  geom_sf(aes(fill = resist_bin), color = "white", linewidth = 0.15) +
  scale_fill_manual(values = c("0-19%"="#FFFFB2","20-39%"="#FECC5C","40-59%"="#FD8D3C",
                                "60-79%"="#F03B20","80-100%"="#BD0026"),
                     name = "% MDR\nE. coli", drop = FALSE) +
  theme_void() + labs(subtitle = "NOT SO GREAT: naive round-number bins")
```

<!--# Real data tops out ~25% — almost every state falls in one bucket. Bins that ignore the actual data range erase your signal. -->

## Much better {.smaller}

Which is better for **your** audience?

::::: columns
::: {.column width="50%"}
**Classed, quantile breaks**

```{r map-great-classed, fig.height=3.2}
qb <- quantile(map_data$percent_resistant, probs = seq(0,1,.2), na.rm = TRUE)
map_data %>%
  mutate(bin = cut(percent_resistant, qb, include.lowest = TRUE,
                    labels = c("Lowest 20%","2nd","3rd","4th","Highest 20%"))) %>%
  ggplot() + geom_sf(aes(fill = bin), color = "white", linewidth = 0.15) +
  scale_fill_brewer(palette = "YlOrRd", name = NULL) + theme_void()
```
:::

::: {.column width="50%"}
**Continuous, red-toned**

```{r map-great-continuous, fig.height=3.2}
ggplot(map_data) +
  geom_sf(aes(fill = percent_resistant), color = "white", linewidth = 0.15) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, name = NULL) +
  theme_void()
```
:::
:::::

<!--# Classed = easier for a general audience to read at a glance. Continuous = full precision for an analyst. Neither is "wrong." -->

## Missing data: Wyoming {.smaller}

```{r drop-missing}
map_data %>% st_drop_geometry() %>% filter(is.na(percent_resistant)) %>% select(name)
map_data <- map_data %>% filter(!is.na(percent_resistant))
```

Wyoming's rate is **suppressed** in the source data — small states often are, to avoid re-identifying patients. This is a real, recurring feature of small-area public health data, not a cleaning mistake. Drop it, or impute deliberately — but say which, and why.

------------------------------------------------------------------------

# Part 5 — Small multiples across years

## Why: a snapshot vs. change {.smaller}

Most interesting environmental-pharmacoepi questions are about *change*. A **small-multiples panel** repeats the same map once per time period so the eye can compare directly.

```{r beam-multiyear}
beam_panel <- expand_grid(stusps = states_conus$stusps, year = 2018:2025) %>%
  left_join(
    beam %>% filter(source_type == "Environment", pathogen == "Salmonella",
                     year >= 2018, year <= 2025, state %in% valid_state_codes) %>%
      group_by(state, year) %>%
      summarize(isolates = sum(number_of_isolates, na.rm = TRUE), .groups = "drop"),
    by = c("stusps" = "state", "year")
  ) %>%
  mutate(isolates = coalesce(isolates, 0))

panel_sf <- states_conus %>% select(stusps, geometry) %>% left_join(beam_panel, by = "stusps")
```

*(We dropped 2026 — partial year would look like a artificial decline.)*

## Map: all years on one scale

```{r facet-years, fig.height=5.2, fig.width=10}

ggplot(panel_sf) +
  geom_sf(aes(fill = isolates), color = "white", linewidth = 0.05) +
  facet_wrap(~ year, ncol = 4) +
  scale_fill_distiller(palette = "YlOrRd",  direction = 1, name = "Environmental\nSalmonella\nisolates",
                       limits = c(0, max(panel_sf$isolates))) +
  theme_void(base_size = 9) +
  theme(strip.text = element_text(face = "bold")) +
  labs(title = "Environmentally-sourced Salmonella isolates by state, 2018-2025",
       subtitle = "CDC BEAM enteric bacteria surveillance",
       caption = "Source: BEAM Dashboard")
```

## Why one shared scale? {.smaller}

Every panel uses **one** color scale built from all 8 years combined.

Map each year with its own default scale instead, and a state with 5 isolates in a low year looks just as "dark" as one with 50 in a high year — same color, wildly different magnitude.

> A shared scale is what makes a small-multiples panel an honest comparison, not eight unrelated pictures that happen to share an axis.

## Discussion {.smaller}

1.  Which state(s) drive most of the signal? Check the individual numbers against your visual read.
2.  Any year jump broadly, across many unrelated states at once? That smells like a **reporting/methods change**, not eight independent outbreaks.
3.  Which states report *consistently* across all 8 years vs. show up once or twice — does that change your read on contamination vs. testing effort?

------------------------------------------------------------------------

# Bonus — Comparing across sources

## Salmonella isolates aren't sampled equally {.smaller}

```{r source-totals, fig.height=3.6}
salm_source_totals <- beam %>%
  filter(pathogen == "Salmonella", year >= 2018, year <= 2025, state %in% valid_state_codes) %>%
  group_by(source_type) %>%
  summarize(isolates = sum(number_of_isolates, na.rm = TRUE), .groups = "drop")

ggplot(salm_source_totals, aes(x = reorder(source_type, isolates), y = isolates)) +
  geom_col(fill = "#440154FF") + coord_flip() +
  scale_y_log10(labels = comma) +
  labs(x = NULL, y = "Total isolates, 2018-2025 (log10 axis)",
       title = "Salmonella isolates by source type") + theme_minimal()
```

Human ≈ 387,000. Environment ≈ 3,000. **Over 100× apart.**

## Map: source type, log scale {.smaller}

```{r source-map, fig.width=10, fig.height=5.6}
source_types <- c("Environment", "Animal", "Food", "Human")
salm_source_sf <- states_conus %>% select(stusps, geometry) %>%
  left_join(
    expand_grid(stusps = states_conus$stusps, source_type = source_types) %>%
      left_join(beam %>% filter(pathogen == "Salmonella", year >= 2018, year <= 2025,
                                 state %in% valid_state_codes) %>%
                  group_by(state, source_type) %>%
                  summarize(isolates = sum(number_of_isolates, na.rm = TRUE), .groups = "drop"),
                by = c("stusps" = "state", "source_type")) %>%
      mutate(isolates = coalesce(isolates, 0), source_type = factor(source_type, levels = source_types)),
    by = "stusps")
```

## Map output

```{r source-map2, fig.width=15, fig.height=3}
ggplot(salm_source_sf) +
  geom_sf(aes(fill = isolates), color = "white", linewidth = 0.05) +
  facet_wrap(~ source_type, nrow = 1) +
  scale_fill_distiller(palette = "YlOrRd",  direction = 1,
                       name = "Salmonella\nisolates\n(log10 scale)",
                       trans = "log1p",
                       breaks = c(0, 10, 100, 1000, 10000)) +
  theme_void(base_size = 9) +
  theme(strip.text = element_text(face = "bold")) +
  labs(title = "Salmonella isolates by source type and state, 2018-2025 pooled",
       subtitle = "CDC BEAM enteric bacteria surveillance",
       caption = "Source: BEAM Dashboard. Color scale is log-transformed -- do not compare panels visually as if linear.")

```

------------------------------------------------------------------------

# Part 6 — Is there a seasonal pattern?

## 6a. Non-spatial first {.smaller}

```{r seasonality-national, fig.height=3.6}
monthly_totals <- beam %>%
  filter(source_type == "Environment", pathogen == "Salmonella",
         year >= 2018, year <= 2025, state %in% valid_state_codes) %>%
  group_by(month) %>% summarize(isolates = sum(number_of_isolates, na.rm = TRUE), .groups = "drop")

ggplot(monthly_totals, aes(x = factor(month), y = isolates)) +
  geom_col(fill = "#21908CFF") + scale_x_discrete(labels = month.abb) +
  labs(x = NULL, y = "Isolates, 2018-2025 pooled", title = "Monthly distribution") +
  theme_minimal()
```

## 6b. Map the same idea

```{r seasonality-spatial, fig.width=10, fig.height=4.2}
season_levels <- c("Winter (Dec-Feb)","Spring (Mar-May)","Summer (Jun-Aug)","Fall (Sep-Nov)")
season_sf <- states_conus %>% select(stusps, geometry) %>%
  left_join(
    expand_grid(stusps = states_conus$stusps, season = season_levels) %>%
      left_join(beam %>% filter(source_type == "Environment", pathogen == "Salmonella",
                                 year >= 2018, year <= 2025, state %in% valid_state_codes) %>%
                  mutate(season = case_when(
                    month %in% c(12,1,2) ~ "Winter (Dec-Feb)", month %in% c(3,4,5) ~ "Spring (Mar-May)",
                    month %in% c(6,7,8) ~ "Summer (Jun-Aug)", month %in% c(9,10,11) ~ "Fall (Sep-Nov)")) %>%
                  group_by(state, season) %>%
                  summarize(isolates = sum(number_of_isolates, na.rm = TRUE), .groups = "drop"),
                by = c("stusps" = "state", "season")) %>%
      mutate(isolates = coalesce(isolates, 0), season = factor(season, levels = season_levels)),
    by = "stusps")
```

## Map output

```{r facet-season-map, fig.width=15, fig.height=3}
ggplot(season_sf) +
  geom_sf(aes(fill = isolates), color = "white", linewidth = 0.05) +
  facet_wrap(~ season, nrow = 1) +
  scale_fill_distiller(palette = "YlOrRd",  direction = 1,
                       name = "Environmental\nSalmonella\nisolates",
                        limits = c(0, max(season_sf$isolates))) +
  theme_void(base_size = 9) +
  theme(strip.text = element_text(face = "bold")) +
  labs(title = "Environmental Salmonella isolates by season, 2018-2025 pooled",
       subtitle = "CDC BEAM enteric bacteria surveillance",
       caption = "Source: BEAM Dashboard")
```

## Discussion {.smaller}

1.  Warm-season *Salmonella* peak is biologically plausible — does the summer panel support it? What competing explanation (surveillance effort) must you rule out first?
2.  Same states lighting up here as in Part 5's year panels? Consistently-high ≠ seasonally-high.
3.  Why pool 8 years into 4 seasons instead of mapping one month of one year? (How sparse would that cell be?)

------------------------------------------------------------------------

# Part 7 — Linking exposure to outcome

## Multi-year signal vs. clinical burden {.smaller}

```{r scatter, fig.height=3.8}
env_total <- beam_panel %>% group_by(stusps) %>%
  summarize(env_isolates_total = sum(isolates), .groups = "drop")

plot_df <- map_data %>% st_drop_geometry() %>% left_join(env_total, by = "stusps")

ggplot(plot_df, aes(x = env_isolates_total, y = percent_resistant)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "firebrick") +
  labs(x = "Environmental Salmonella isolates, 2018-2025", y = "% MDR E. coli, 2024") +
  theme_minimal()

cor.test(plot_df$env_isolates_total, plot_df$percent_resistant, method = "spearman")
```

## Two concepts relevant to these findings {.smaller}

**Ecological fallacy** — even a state-level correlation tells you nothing about an *individual* patient's exposure-outcome link.

**MAUP** (Modifiable Areal Unit Problem) — aggregate to counties, HHS regions, or watersheds instead of states, and both the magnitude *and direction* could change, with zero change in underlying reality.

> If you designed a real study on this question: what exposure variable, at what spatial *and temporal* resolution, would you actually want?

------------------------------------------------------------------------

# Wrap-up

## Carry these questions out of the room {.smaller}

- Everything today was area-level, pre-aggregated data. What changes with **point-level** data (addresses, wells, facilities)? Think buffers, distance-to-source, kernel density.
- Where do *you* use a "zone" (ZIP, county, tract, state) just because that's the unit your data comes in — not because it's the right scale for the exposure mechanism?
- Surveillance-effort bias showed up clearly today. Where might it be hiding, less visibly, in a dataset you already use?

## Where to go next {.smaller}

Today stopped at "here's the pattern, here's why to be cautious." The natural next question — is a visible pattern *actually* clustered more than chance? — is spatial autocorrelation (Moran's I / LISA, via `spdep`). Self-study follow-on, not covered today.

**Further resources**

- Pebesma & Bivand, *Spatial Data Science* (free online)
- `tigris` docs — other Census geographies (counties, ZCTAs, tracts)
- `spdep` vignettes, for the follow-on exercise
