---
title: "ANSWER KEY: Mapping Antimicrobial Resistance Across U.S. States"
subtitle: "An Introduction to GIS Methods for Environmental Pharmacoepidemiology"
author: "ISPE 2026 Pre-Conference Course"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5.5)
```

> **Depends on the main exercise.** This file's answer-key chunks
> (`answer-key-panel`, `answer-key-season`, etc.) reuse `beam`,
> `states_conus`, and `valid_state_codes` from
> `exercises/GIS_Intro_Exercise_EnvPharmacoepi.Rmd` — they aren't
> reloaded here. Run the main exercise first in the same R session (or
> `source()` it) before running or knitting this file on its own.

------------------------------------------------------------------------

# Part 8 — Your turn: independent practice

Two parts here. First, a quick recap of the join-and-map workflow on a
new file. Second, and more importantly, you'll build your own
small-multiples panels — the core new skill from today — on a different
slice of the BEAM data.

## 8.1 Quick recap: import and map the MRSA SIR data

Use **Table 7 (state-level hospital-onset MRSA bacteremia SIRs)** from
`2024-SIR-ACH.xlsx`.

The sheet has a multi-row title block above the real header row — this
is extremely common in Excel reports designed to be read by humans, not
software. Use `readxl::read_excel()` and skip past the title rows so
that the row containing "State", "SIR", etc. becomes your header.

```{r your-turn-import, eval=FALSE}
mrsa_raw <- read_excel("../data/2024-SIR-ACH.xlsx",
                        sheet = "Table 7-State MRSA Data",
                        skip = ____)   # how many rows do you need to skip?

mrsa <- mrsa_raw %>%
  clean_names() %>%
  select(state, sir) %>%
  mutate(sir = suppressWarnings(as.numeric(sir))) %>%
  filter(!is.na(state), state != "____")   # there's a US-total summary row at the bottom -- what is it called?
```

**Look closely at `unique(mrsa$state)` before you join.** Different CDC
data products don't always spell the same state the same way — this
table uses `"D.C."`, while `tigris` (and the ARPSP file you used
earlier) use `"District of Columbia"`. A join is only as good as its
key, so fix any mismatches you find *before* joining, e.g.:

```{r your-turn-recode, eval=FALSE}
mrsa <- mrsa %>%
  mutate(state = recode(state, "____" = "District of Columbia"))
```

```{r your-turn-join, eval=FALSE}
mrsa_map <- states_conus %>%
  left_join(mrsa, by = c("____" = "____"))   # which key matches full state names?
```

Check the join the same way we did in Part 3 — did any states fail to
match? If `sir` still has an `NA` in it once the join looks clean, that's
not a join problem — it's a **suppressed value**, exactly like Wyoming
in Part 4. Decide, and state explicitly, whether you'd drop or impute
it.

```{r your-turn-map, eval=FALSE}
ggplot(mrsa_map) +
  geom_sf(aes(fill = ____)) +
  scale_fill_viridis_c(name = "MRSA SIR", option = "____", direction = ____) +
  theme_void() +
  labs(title = "____")
```

## 8.2 Build your own small-multiples panel

Pick a **pathogen** and a **source type** from the BEAM data that we
*haven't* used yet — anything other than Environment + Salmonella. Try
`unique(beam$pathogen)` and `unique(beam$source_type)` if you want
options. Food-sourced Campylobacter and Animal-sourced Salmonella are
both good choices with plenty of data behind them.

**Step 1 — a year panel**, following the exact pattern from Part 5:

```{r your-turn-panel, eval=FALSE}
my_pathogen    <- "____"
my_source_type <- "____"

my_panel <- beam %>%
  filter(source_type == my_source_type, pathogen == my_pathogen,
         year >= 2018, year <= 2025,
         state %in% valid_state_codes) %>%
  group_by(state, ____) %>%                          # group by what, this time?
  summarize(isolates = sum(number_of_isolates, na.rm = TRUE), .groups = "drop")

my_grid <- expand_grid(stusps = states_conus$stusps, year = 2018:2025) %>%
  left_join(my_panel, by = c("stusps" = "state", "year")) %>%
  mutate(isolates = coalesce(isolates, ____))         # fill missing with what?

my_panel_sf <- states_conus %>%
  select(stusps, geometry) %>%
  left_join(my_grid, by = "stusps")

ggplot(my_panel_sf) +
  geom_sf(aes(fill = ____), color = "white", linewidth = 0.05) +
  facet_wrap(~ ____, ncol = 4) +
  scale_fill_viridis_c(limits = c(0, max(my_panel_sf$isolates))) +
  theme_void() +
  labs(title = "____")
```

**Step 2 — a season panel**, following the exact pattern from Part 6.
Reuse the `case_when()` month groupings from Part 6b, then build the
grid, join, and facet map yourself from scratch — you have the full
pattern twice now (year panel above, season panel in Part 6).

## 8.3 Interpret

Answer in a sentence or two each:

- Compared to our Environment + Salmonella example, does your
  pathogen/source combination look *denser* (more states, more isolates
  per state) or *sparser*? Which source types in BEAM are sampled far
  more heavily than others, and what does that imply about comparing
  "environment" numbers to "food" or "animal" numbers head to head?
- Does your panel show a clearer year-to-year trend, a clearer seasonal
  pattern, both, or neither? What's one non-biological explanation you'd
  want to rule out before trusting either pattern?
- If a colleague showed you one of these small-multiples panels without
  telling you it used a shared color scale, and you found out later it
  hadn't — how would that change what conclusions you'd trust from it?

------------------------------------------------------------------------
------------------------------------------------------------------------

```{=html}
<!-- =====================================================================
ANSWER KEY -- Part 8
===================================================================== -->
```

# Instructor answer key (Part 8)

## 8.1 answer key

```{r answer-key-import}
mrsa_raw <- read_excel("../data/2024-SIR-ACH.xlsx",
                        sheet = "Table 7-State MRSA Data",
                        skip = 4)

mrsa <- mrsa_raw %>%
  clean_names() %>%
  select(state, sir) %>%
  mutate(sir = suppressWarnings(as.numeric(sir))) %>%
  filter(!is.na(state), state != "All US") %>%
  # tigris/ARPSP use "District of Columbia"; this sheet uses "D.C." -- fix
  # the key before joining, or DC will silently fail to match.
  mutate(state = recode(state, "D.C." = "District of Columbia"))
```

```{r answer-key-join}
mrsa_map <- states_conus %>%
  left_join(mrsa, by = c("name" = "state"))

mrsa_map %>% st_drop_geometry() %>% filter(is.na(sir)) %>% select(name)
```

Guam and the Virgin Islands report `sir = "."` (suppressed) in the
source sheet, but since they aren't part of `states_conus` to begin
with, they never enter `mrsa_map` — nothing to clean up there. The join
check above should come back empty once the DC recode is in place.

```{r answer-key-map}
ggplot(mrsa_map) +
  geom_sf(aes(fill = sir), color = "white", linewidth = 0.15) +
  scale_fill_viridis_c(name = "MRSA SIR", option = "magma", direction = -1) +
  theme_void() +
  labs(title = "Hospital-onset MRSA bacteremia SIR by state, 2024 NHSN")
```

## 8.2 answer key (worked with Campylobacter, Food source)

```{r answer-key-panel, fig.width=10, fig.height=6}
my_pathogen    <- "Campylobacter"
my_source_type <- "Food"

my_panel <- beam %>%
  filter(source_type == my_source_type, pathogen == my_pathogen,
         year >= 2018, year <= 2025,
         state %in% valid_state_codes) %>%
  group_by(state, year) %>%
  summarize(isolates = sum(number_of_isolates, na.rm = TRUE), .groups = "drop")

my_grid <- expand_grid(stusps = states_conus$stusps, year = 2018:2025) %>%
  left_join(my_panel, by = c("stusps" = "state", "year")) %>%
  mutate(isolates = coalesce(isolates, 0))

my_panel_sf <- states_conus %>%
  select(stusps, geometry) %>%
  left_join(my_grid, by = "stusps")

ggplot(my_panel_sf) +
  geom_sf(aes(fill = isolates), color = "white", linewidth = 0.05) +
  facet_wrap(~ year, ncol = 4) +
  scale_fill_viridis_c(name = "Food-sourced\nCampylobacter\nisolates",
                        limits = c(0, max(my_panel_sf$isolates))) +
  theme_void(base_size = 9) +
  theme(strip.text = element_text(face = "bold")) +
  labs(title = "Food-sourced Campylobacter isolates by state, 2018-2025")
```

Food-sourced Campylobacter is a good contrast case: roughly 45-49 states
report in every single year (vs. ~14-26 for Environment-sourced
Salmonella), and annual totals run in the thousands rather than the
hundreds. That density is exactly why food and animal surveillance
streams are more often used as primary exposure variables in published
work than environmental streams — not because the environmental pathway
matters less biologically, but because the data are thinner.

```{r answer-key-season, fig.width=10, fig.height=4}
season_levels <- c("Winter (Dec-Feb)", "Spring (Mar-May)",
                    "Summer (Jun-Aug)", "Fall (Sep-Nov)")

my_season_data <- beam %>%
  filter(source_type == my_source_type, pathogen == my_pathogen,
         year >= 2018, year <= 2025,
         state %in% valid_state_codes) %>%
  mutate(season = case_when(
    month %in% c(12, 1, 2)  ~ "Winter (Dec-Feb)",
    month %in% c(3, 4, 5)   ~ "Spring (Mar-May)",
    month %in% c(6, 7, 8)   ~ "Summer (Jun-Aug)",
    month %in% c(9, 10, 11) ~ "Fall (Sep-Nov)"
  )) %>%
  group_by(state, season) %>%
  summarize(isolates = sum(number_of_isolates, na.rm = TRUE), .groups = "drop")

my_season_panel <- expand_grid(stusps = states_conus$stusps, season = season_levels) %>%
  left_join(my_season_data, by = c("stusps" = "state", "season")) %>%
  mutate(isolates = coalesce(isolates, 0),
         season = factor(season, levels = season_levels))

my_season_sf <- states_conus %>%
  select(stusps, geometry) %>%
  left_join(my_season_panel, by = "stusps")

ggplot(my_season_sf) +
  geom_sf(aes(fill = isolates), color = "white", linewidth = 0.05) +
  facet_wrap(~ season, nrow = 1) +
  scale_fill_viridis_c(name = "Food-sourced\nCampylobacter\nisolates",
                        limits = c(0, max(my_season_sf$isolates))) +
  theme_void(base_size = 9) +
  theme(strip.text = element_text(face = "bold")) +
  labs(title = "Food-sourced Campylobacter isolates by season, 2018-2025 pooled")
```

National monthly totals for this combination run from about 2,250 (Feb)
to about 3,240 (Jun) — a modest but real-looking early-summer bump,
plausible for a foodborne pathogen tied to warm-weather food handling
and outdoor eating, and visibly less driven by a handful of outlier
states than the Environment + Salmonella example, precisely because the
underlying counts are so much larger.

**Interpretation talking points for 8.3:** the biggest single lesson to
draw out here is that BEAM's four source types (Animal, Food, Human,
Environment) are not equally sampled, and a denser stream will always
look "smoother" and more confidently seasonal than a sparse one even if
the true underlying biology were identical — surveillance intensity is a
feature of the data-generating process, not of the exposure itself, and
it has to be reasoned about explicitly rather than read off the map.
