---
title: "Mapping Antimicrobial Resistance Across U.S. States"
subtitle: "An Introduction to GIS Methods for Environmental Pharmacoepidemiology"
author: "ISPE 2026 Pre-Conference Course"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5.5)
```


# Welcome

We are going to build a complete GIS workflow in R, using real
CDC surveillance data:

1.  **`ARPSP_MDR-Ecoli---Percent-Resistant-in-2024.csv`** — the percent
    of multidrug-resistant (MDR) *E. coli* isolates among
    healthcare-associated infections, by state, from CDC's Antibiotic
    Resistance & Patient Safety Portal (ARPSP), 2024. 
2.  **`BEAM_Dashboard_-_Report_Data_20260804.csv`** — CDC enteric
    bacteria surveillance isolates (Salmonella, *E. coli*/STEC,
    Campylobacter, Shigella, Vibrio) drawn from animal, food, human, and
    **environmental** sources, reported monthly since 2018. We will use
    the environmental-source *Salmonella* records as a proxy **exposure
    / surveillance-effort**.
3.  **`2024-SIR-ACH.xlsx`** — CDC NHSN Standardized Infection Ratio
    (SIR) report for acute care hospitals, 2024. You will use this
    yourself in Part 8.

None of these files contain a shapefile, a latitude, or a longitude.
That is deliberate. Most pharmacoepidemiology environmental exposure
data you will encounter in practice — surveillance reports, claims data
aggregated for privacy, vital statistics — arrives as **tables tied to
an administrative area** (state, county, ZIP) **and to a time period**
(month, year).

## Learning objectives

By the end, you will be able to:

- Import and clean real, messy public-health surveillance data in R
- Obtain U.S. state boundary polygons and understand what a coordinate
  reference system (CRS) is doing for you
- Perform an **attribute join** — attaching non-spatial data to spatial
  geometry by a shared key — the single most common GIS operation you
  will do in this field
- Build choropleth (thematic) maps with `ggplot2` + `sf`
- Build **small-multiples map panels** to show how a distribution
  changes across years and across seasons, and understand why the color
  scale has to be shared across panels for the comparison to be fair
- Recognize **seasonality** in area-level surveillance data using both a
  simple distribution plot and a mapped version of the same pattern
- Name the specific ways area-level GIS analysis can mislead you in
  environmental pharmacoepidemiology (MAUP, ecological fallacy,
  surveillance effort bias) — and recognize them in your own results

------------------------------------------------------------------------

# Setup

```{r install-packages, eval=FALSE}
# Run once. Skip if already installed.
install.packages(c("sf", "tigris", "dplyr", "tidyr", "readr", "readxl",
                    "janitor", "ggplot2", "viridis", "scales"))
```

```{r load-packages}
library(sf)        # spatial vector data: geometry + attributes
library(tigris)     # download US Census boundary files
library(dplyr)       # data wrangling
library(tidyr)         # expand_grid(), for building complete state x time panels
library(readr)          # fast, sane CSV reading
library(readxl)           # read Excel sheets
library(janitor)           # clean up ugly column names
library(ggplot2)             # mapping and plotting
library(viridis)               # colorblind-safe palettes
library(scales)                  # nicer plot labels

options(tigris_use_cache = TRUE)   # cache downloaded shapefiles locally
sf::sf_use_s2(TRUE)                 # spherical geometry engine (default, just being explicit)
```

> **Teaching note:** `tigris` downloads boundary files from the U.S.
> Census Bureau the first time you use them, then caches them on disk.
> If you are offline, use the pre-downloaded
> `states_sf.rds` file and load it with
> `states_sf <- readRDS("states_sf.rds")` instead of running the
> `tigris::states()` call below.

------------------------------------------------------------------------

# Part 1 — Import and clean the tabular data

## 1a. Clinical outcome: MDR *E. coli* resistance by state

```{r import-arpsp}
ecoli_raw <- read_csv("../data/ARPSP_MDR-Ecoli---Percent-Resistant-in-2024.csv")

ecoli <- ecoli_raw %>%
  clean_names() %>%                 # tidies "Percent Resistant" -> percent_resistant
  select(state, percent_resistant) %>%
  mutate(percent_resistant = as.numeric(percent_resistant))

glimpse(ecoli)
```

**Stop and look.** How many rows do you have? Is that one row per U.S.
state? (Hint: `nrow(ecoli)`; check whether Puerto Rico and DC are
included with `unique(ecoli$state)`.) Knowing your unit of observation
*before* you map it is a GIS habit, as it will help you
determine what your eventual polygon layer needs to contain. 

## 1b. Environmental exposure signal: BEAM environmental isolates by state

The raw BEAM file is semicolon-delimited and has a European-style
thousands separator in the `Year` column (e.g. `"2,024"`). This is
completely typical of real surveillance exports.

```{r import-beam}
beam_raw <- read_delim("../data/BEAM_Dashboard_-_Report_Data_20260804.csv",
                        delim = ";", show_col_types = FALSE)

beam <- beam_raw %>%
  clean_names() %>%
  mutate(year = as.numeric(gsub(",", "", year)))

# Keep only environmental-source Salmonella isolates from 2024, as a
# first, simple look -- one state, one year, one map.
beam_env_2024 <- beam %>%
  filter(year == 2024, source_type == "Environment", pathogen == "Salmonella")

# The "state" column has some junk values in it (data-entry artifacts like
# "N/", "US", "OU"). A real GIS key needs to be *valid* before you can join
# on it -- always sanity-check your join key.
valid_state_codes <- c(state.abb, "DC")   # state.abb is a built-in R constant
sort(unique(beam_env_2024$state))
```

```{r clean-beam}
beam_env_clean <- beam_env_2024 %>%
  filter(state %in% valid_state_codes) %>%
  group_by(state) %>%
  summarize(env_isolates = sum(number_of_isolates, na.rm = TRUE), .groups = "drop")

beam_env_clean
```

**Discussion:** Only about 20 states show up here at all, for 2024 alone.
Is that because *Salmonella* genuinely isn't present in the environment
elsewhere — or because those states simply aren't sampling environmental
sources? Hold onto that question; we'll come back to it, with more data
behind us, in Parts 5-7. In environmental pharmacoepidemiology, an
"exposure" layer built from surveillance data is never just measuring
exposure — it's also measuring *where people looked.* 

**Sampling choices**


------------------------------------------------------------------------

# Part 2 — Get the map: state boundary polygons

```{r get-states}
states_sf <- tigris::states(cb = TRUE, resolution = "20m", year = 2025)

states_sf <- states_sf %>%
  clean_names() %>%
  select(name, stusps, geoid, geometry)

glimpse(states_sf)
```

`states_sf` is an `sf` object: an ordinary data frame with one extra
column, `geometry`, holding the polygon boundary for each state.
Everything you already know about `dplyr::filter()`, `mutate()`, and
`left_join()` still works!!.
 
Offline fallback: if tigris::states() times out or you have no internet,
download the cb_2025_us_state_20m folder next to this .Rmd (or adjust the
path below), then use this instead of the tigris::states() call.
  states_sf <- sf::read_sf("cb_2025_us_state_20m/cb_2025_us_state_20m.shp")

`states_sf <- states_sf %>%
  clean_names() %>%
  select(name, stusps, geoid, geometry)`

`glimpse(states_sf)`

```{r plot-blank-map}
ggplot(states_sf) +
  geom_sf() +
  theme_minimal() +
  labs(title = "U.S. state boundaries (no data joined yet)")
```

For the rest of this exercise we'll restrict to the contiguous 48
states + DC. Alaska, Hawaii, Puerto Rico, and the outlying territories
are real and important, but they sit far outside the CONUS bounding box
and would either get cut off or force us to shrink every other state to
fit them on the same simple map. (`tigris::shift_geometry()` can
reposition them into inset boxes if you need a true all-50-states map
later — worth knowing exists, out of scope for today.) For a first pass
at small-multiples mapping, keeping the geography simple lets us focus
on the actual skill: comparing panels to each other.

```{r conus-only}
conus_codes <- setdiff(c(state.abb, "DC"), c("AK", "HI"))

states_conus <- states_sf %>%
  filter(stusps %in% conus_codes) %>%
  st_transform(5070)   # Albers Equal Area Conic — standard projected CRS for CONUS
```

> **CRS check-in.** `tigris` gives you data in geographic coordinates
> (NAD83, unprojected — degrees of latitude/longitude). We just
> projected to EPSG:5070, an *equal-area* projection built for the
> contiguous U.S. Projection matters here for two concrete reasons: (1)
> `ggplot2` will draw a nicer, less distorted map, and (2) any analysis
> that depends on area or distance (which ours doesn't today, but often
> will) requires a projected CRS to be correct. Always ask "what CRS is
> this in, and does that CRS make sense for what I'm about to compute?"

------------------------------------------------------------------------

# Part 3 — Attribute join

This is the step that turns a spreadsheet into a map. We attach our two
non-spatial tables to `states_conus` by matching a shared key: full
state name for the ARPSP data, two-letter abbreviation for the BEAM
data.

```{r attribute-join}
map_data <- states_conus %>%
  left_join(ecoli, by = c("name" = "state")) %>%
  left_join(beam_env_clean, by = c("stusps" = "state")) %>%
  mutate(env_isolates = coalesce(env_isolates, 0))  # no report = 0 isolates, not missing

summary(map_data$percent_resistant)
summary(map_data$env_isolates)
```

**Stop and check your join.** A silent join failure — rows that should
have matched but didn't, because of a trailing space, a "St." vs
"Saint", a different case — is the single most common source of a blank
or wrong-looking choropleth map. Before you ever trust a joined map,
run:

```{r check-join}
map_data %>%
  st_drop_geometry() %>%
  filter(is.na(percent_resistant)) %>%
  select(name, stusps)
```

If that returns states you *expected* to have data, go back and find the
mismatch. If it's empty, your join key worked.

------------------------------------------------------------------------

# Part 4 — Your first choropleth maps

```{r choropleth-ecoli}
# MAP 1 -- BAD: a qualitative (non-ordered) palette forced onto ordered data.
# Set1's colors have no natural low-to-high sequence -- red isn't obviously
# "more resistant" than blue or green, so the reader has to keep re-checking
# the legend instead of reading the map at a glance. Also rough for
# colorblind readers (red/green together).
map_data_bad <- map_data %>%
  mutate(resist_bin = cut(percent_resistant,
                           breaks = c(0, 5, 10, 15, 25),
                           labels = c("0-4%", "5-9%", "10-14%", "15%+"),
                           include.lowest = TRUE))

ggplot(map_data_bad) +
  geom_sf(aes(fill = resist_bin), color = "white", linewidth = 0.15) +
  scale_fill_brewer(palette = "Set1", name = "% MDR\nE. coli") +
  theme_void() +
  labs(title = "Multidrug-resistant E. coli among HAIs, by state, 2024",
       subtitle = "BAD: a qualitative palette on ordered data",
       caption = "Source: ARPSP")

 
# MAP 2 -- NOT SO GREAT: right idea (sequential yellow-to-red), wrong bins.
# These are naive round-number "percent" buckets (0-19, 20-39, ...) instead
# of bins built from the data's actual range. Real values only go up to
# ~25%, so almost every state falls into the first bucket and the map goes
# nearly flat -- you lose almost all the variation you're trying to show.
map_data_meh <- map_data %>%
  mutate(resist_bin = cut(percent_resistant,
                           breaks = c(0, 20, 40, 60, 80, 100),
                           labels = c("0-10%", "10-19%", "20-39%", "40-59%", "60-100%"),
                           include.lowest = TRUE))

ggplot(map_data_meh) +
  geom_sf(aes(fill = resist_bin), color = "white", linewidth = 0.15) +
  scale_fill_manual(values = c("0-10%" = "#FFFFB2", "10-19%" = "#FECC5C",
                                "20-39%" = "#FD8D3C", "40-59%" = "#F03B20",
                                "60-100%" = "#BD0026"),
                     name = "% MDR\nE. coli", drop = FALSE) +
  theme_void() +
  labs(title = "Multidrug-resistant E. coli among HAIs, by state, 2024",
       subtitle = "NOT SO GREAT: sequential color, but bins don't match the data",
       caption = "Source: ARPSP")

# GREAT, classed version: quantile breaks so every bin has roughly the same
# number of states (no bin swallows the whole map, unlike Map 2's round
# numbers) + a red-toned sequential palette.
quantile_breaks <- quantile(map_data$percent_resistant, probs = seq(0, 1, 0.2), na.rm = TRUE)

map_data_great <- map_data %>%
  mutate(resist_bin = cut(percent_resistant, breaks = quantile_breaks,
                           include.lowest = TRUE,
                           labels = c("Lowest 20%", "2nd", "3rd", "4th", "Highest 20%")))

ggplot(map_data_great) +
  geom_sf(aes(fill = resist_bin), color = "white", linewidth = 0.15) +
  scale_fill_brewer(palette = "YlOrRd", name = "% MDR\nE. coli\n(quintile)") +
  theme_void() +
  labs(title = "Multidrug-resistant E. coli among HAIs, by state, 2024",
       subtitle = "GREAT: classed, quantile breaks, red = higher resistance",
       caption = "Source: ARPSP")


ggplot(map_data) +
  geom_sf(aes(fill = percent_resistant), color = "white", linewidth = 0.15) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "% MDR\nE. coli") +
  theme_void() +
  labs(title = "Multidrug-resistant E. coli among HAIs, by state, 2024",
       subtitle = "GREAT: continuous, red = higher resistance",
       caption = "Source: ARPSP")

```

```{r choropleth-beam}


ggplot(map_data) +
  geom_sf(aes(fill = env_isolates), color = "white", linewidth = 0.15) +
  scale_fill_viridis_c(name = "Environmental\nSalmonella\nisolates", option = "viridis") +
  theme_void() +
  labs(title = "Environmentally-sourced Salmonella isolates, by state, 2024",
       subtitle = "CDC BEAM enteric bacteria surveillance",
       caption = "Source: BEAM Dashboard")

ggplot(map_data) +
  geom_sf(aes(fill = env_isolates), color = "white", linewidth = 0.15) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "% MDR\nSalmonella") +
  theme_void() +
  labs(title = "Environmentally-sourced Salmonella isolates, by state, 2024",
       subtitle = "CDC BEAM enteric bacteria surveillance",
       caption = "Source: BEAM Dashboard")

```

**Discussion prompts:**

1.  `scale_fill_viridis_c()` uses a *continuous* color gradient. What
    would change if you binned this into 4-5 categories instead
    (`scale_fill_viridis_b()`)? What decision are you making implicitly
    when you choose bin breaks (equal interval vs. quantile vs. Jenks
    natural breaks)? This is not a cosmetic choice — bin choice can
    visually create or hide a "hot spot."
2.  Look at the second map. Does it look like a map of environmental
    *Salmonella* contamination, or a map of *where states happened to
    test environmental samples*? 

## A note on missing data before we go further

Run `summary(map_data$percent_resistant)` again and look for `NA's`. One
state — Wyoming — has a missing value in the ARPSP source data
(not a join error). Small states often have their rates **suppressed**
in public surveillance data when the underlying case count is too small
to report reliably without a risk of re-identifying a patient. This is a
real and recurring feature of small-area public health data, not a
data-cleaning mistake to hide.

```{r drop-missing}
map_data %>% st_drop_geometry() %>% filter(is.na(percent_resistant)) %>% select(name)

map_data <- map_data %>% filter(!is.na(percent_resistant))
nrow(map_data)   # 48 states remain
```

**Discussion:** dropping is not the only option — you could instead
impute Wyoming's value (e.g., a regional mean), but imputation should be
a deliberate, documented choice.

------------------------------------------------------------------------

# Part 5 — Expanding across years: small multiples of environmental surveillance

A single choropleth is a snapshot. Most of the interesting questions in
environmental pharmacoepidemiology are about *change* — did something
increase, decrease, or stay flat? The standard way to show that
spatially is a **small-multiples panel**: the same map, repeated once
per time period, laid out in a grid so the eye can compare them directly.

Let's rebuild the BEAM environmental *Salmonella* extract, this time
keeping every full year from 2018 through 2025 (we'll drop 2026 — it's
only partially reported as of today, and a partial year next to eight
full years would make the most recent panel look like a decline that
isn't real).

```{r beam-multiyear}
beam_env_years <- beam %>%
  filter(source_type == "Environment", pathogen == "Salmonella",
         year >= 2018, year <= 2025,
         state %in% valid_state_codes) %>%
  group_by(state, year) %>%
  summarize(isolates = sum(number_of_isolates, na.rm = TRUE), .groups = "drop")
```

Every CONUS state should get one row for *every* year, even the years it
reported nothing — an absent row and a reported zero mean different
things once you're building a panel, and only one of them is correct
here.

```{r beam-panel-grid}
beam_panel <- expand_grid(
  stusps = states_conus$stusps,
  year   = 2018:2025
) %>%
  left_join(beam_env_years, by = c("stusps" = "state", "year")) %>%
  mutate(isolates = coalesce(isolates, 0))

nrow(beam_panel)   # 49 states x 8 years
```

```{r panel-join-map}
panel_sf <- states_conus %>%
  select(stusps, name, geometry) %>%
  left_join(beam_panel, by = "stusps")
```

```{r facet-years, fig.width=10, fig.height=6}
ggplot(panel_sf) +
  geom_sf(aes(fill = isolates), color = "white", linewidth = 0.05) +
  facet_wrap(~ year, ncol = 4) +
  scale_fill_distiller(palette = "YlOrRd",  direction = 1, name = "Environmental\nSalmonella\nisolates",
                        limits = c(0, max(panel_sf$isolates))) +
  theme_void(base_size = 9) +
  theme(strip.text = element_text(face = "bold")) +
  labs(title = "Environmentally-sourced Salmonella isolates by state, 2018-2025",
       subtitle = "CDC BEAM enteric bacteria surveillance",
       caption = "Source: BEAM Dashboard")
```

**Why the `limits = c(0, max(panel_sf$isolates))`** Every panel
here shares one color scale, built from the range of *all eight years
combined*. If you instead mapped each year separately with its own
default scale, a state with 5 isolates in a low year could look just as
"dark" as a state with 50 isolates in a high year — same color, wildly
different magnitude. A shared scale is what makes a small-multiples
panel an honest comparison instead of eight unrelated pictures that
happen to share an axis. This is arguably the single most common way
choropleth series mislead a reader, and it's a one-line fix.

**Discussion prompts:**

1.  Which one or two states look like they're driving most of the
    signal across the panel? Pull up their individual year-by-year
    isolate counts and check your visual read against the numbers.
2.  Do any years look like a sharp jump or drop for many states at once
    (rather than one state)? A change that shows up broadly, in the same
    year, across unrelated states, is more likely a **reporting or
    methods change** than eight independent outbreaks.
3.  Go back to your answer from Part 4, question 2. Now that you can see
    which states report *consistently* across all eight years versus
    which show up once or twice — does that change your read on whether
    this layer measures contamination, testing effort, or some mix of
    both?

# Total Salmonella isolates by source type, 2018-2025 pooled -- look at the
# scale of the imbalance before you try to map it.
```{r facet}
salm_source_totals <- beam %>%
  filter(pathogen == "Salmonella", year >= 2018, year <= 2025,
         state %in% valid_state_codes) %>%
  group_by(source_type) %>%
  summarize(isolates = sum(number_of_isolates, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(isolates))

salm_source_totals

ggplot(salm_source_totals, aes(x = reorder(source_type, isolates), y = isolates)) +
  geom_col(fill = "#440154FF") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Total Salmonella isolates, 2018-2025 (log10 axis)",
       title = "Salmonella isolates by source type") +
  scale_y_log10(labels = comma) +
  theme_minimal()
```

# State x source-type panel, same zero-fill logic as Parts 5-6.
```{r facet2}
salm_by_source <- beam %>%
  filter(pathogen == "Salmonella", year >= 2018, year <= 2025,
         state %in% valid_state_codes) %>%
  group_by(state, source_type) %>%
  summarize(isolates = sum(number_of_isolates, na.rm = TRUE), .groups = "drop")

source_types <- c("Environment", "Animal", "Food", "Human")

salm_source_panel <- expand_grid(stusps = states_conus$stusps, source_type = source_types) %>%
  left_join(salm_by_source, by = c("stusps" = "state", "source_type")) %>%
  mutate(isolates = coalesce(isolates, 0),
         source_type = factor(source_type, levels = source_types))

salm_source_sf <- states_conus %>%
  select(stusps, geometry) %>%
  left_join(salm_source_panel, by = "stusps")

# Raw counts on one shared scale would make Environment/Animal/Food look
# essentially blank next to Human -- log10(isolates + 1) lets all four
# panels show internal spatial variation at once. Note in the legend that
# it's a log scale, NOT LINEAR
ggplot(salm_source_sf) +
  geom_sf(aes(fill = isolates), color = "white", linewidth = 0.05) +
  facet_wrap(~ source_type, nrow = 1) +
  scale_fill_distiller(palette = "YlOrRd",  direction = 1,
                       name = "Salmonella\nisolates\n(log10 scale)",
                        trans = "log1p",
                        breaks = c(0, 10, 100, 1000, 10000)) +
  theme_void(base_size = 9) +
  theme(strip.text = element_text(face = "bold")) +
  labs(title = "Salmonella isolates by source type and state, 2018-2025 pooled",
       subtitle = "CDC BEAM enteric bacteria surveillance",
       caption = "Source: BEAM Dashboard. Color scale is log-transformed -- do not compare panels visually as if linear.")
```
------------------------------------------------------------------------

# Part 6 — Is there a seasonal pattern?

Small multiples don't have to be organized by year — the same technique
works for any repeating unit of time. Here we'll look at *season*, using
all eight years pooled together so each panel has enough data behind it
to be worth looking at.

## 6a. Start non-spatial: what does the distribution look like?

Before mapping anything, look at the plain distribution. Pooling across
2018-2025 gives us enough isolates per month to see a shape, where any
single year would mostly be noise.

```{r seasonality-national}
monthly_totals <- beam %>%
  filter(source_type == "Environment", pathogen == "Salmonella",
         year >= 2018, year <= 2025,
         state %in% valid_state_codes) %>%
  group_by(month) %>%
  summarize(isolates = sum(number_of_isolates, na.rm = TRUE), .groups = "drop")

ggplot(monthly_totals, aes(x = factor(month), y = isolates)) +
  geom_col(fill = "#21908CFF") +
  scale_x_discrete(labels = month.abb) +
  labs(x = NULL, y = "Total isolates, 2018-2025 pooled (all CONUS states)",
       title = "Monthly distribution of environmental Salmonella isolates") +
  theme_minimal()
```

**Discussion:** Does this look seasonal to you, or does it look close to
flat with some noise? What would make you more confident either way —
more years, more states, or something else?

## 6b. Now map the same idea

Group the same eight years of data into four seasons per state, so we
can see whether the pattern in 6a is broadly shared across the country
or driven by a handful of states.

```{r seasonality-spatial}
season_levels <- c("Winter (Dec-Feb)", "Spring (Mar-May)",
                    "Summer (Jun-Aug)", "Fall (Sep-Nov)")

beam_season <- beam %>%
  filter(source_type == "Environment", pathogen == "Salmonella",
         year >= 2018, year <= 2025,
         state %in% valid_state_codes) %>%
  mutate(season = case_when(
    month %in% c(12, 1, 2)  ~ "Winter (Dec-Feb)",
    month %in% c(3, 4, 5)   ~ "Spring (Mar-May)",
    month %in% c(6, 7, 8)   ~ "Summer (Jun-Aug)",
    month %in% c(9, 10, 11) ~ "Fall (Sep-Nov)"
  )) %>%
  group_by(state, season) %>%
  summarize(isolates = sum(number_of_isolates, na.rm = TRUE), .groups = "drop")

season_panel <- expand_grid(stusps = states_conus$stusps, season = season_levels) %>%
  left_join(beam_season, by = c("stusps" = "state", "season")) %>%
  mutate(isolates = coalesce(isolates, 0),
         season = factor(season, levels = season_levels))

season_sf <- states_conus %>%
  select(stusps, geometry) %>%
  left_join(season_panel, by = "stusps")
```

```{r facet-season-map, fig.width=10, fig.height=4}
ggplot(season_sf) +
  geom_sf(aes(fill = isolates), color = "white", linewidth = 0.05) +
  facet_wrap(~ season, nrow = 1) +
  scale_fill_distiller(palette = "YlOrRd",  direction = 1,
                       name = "Environmental\nSalmonella\nisolates",
                        limits = c(0, max(season_sf$isolates))) +
  theme_void(base_size = 9) +
  theme(strip.text = element_text(face = "bold")) +
  labs(title = "Environmental Salmonella isolates by season, 2018-2025 pooled",
       subtitle = "CDC BEAM enteric bacteria surveillance",
       caption = "Source: BEAM Dashboard")
```

**Discussion prompts:**

1.  *Salmonella* survival and growth in water and soil is generally
    understood to be temperature-dependent, which would predict a
    warm-season peak. Does the summer panel support that here? Before
    you answer yes, go back to the surveillance-effort caution from Part
    1b and Part 5 — what competing explanation do you need to rule out
    first?
2.  Are the same states lighting up in Part 5's year panels and this
    season panel? A state that's consistently high across all seasons is
    telling you a different story than a state that's high in exactly
    one season.
3.  Why did we pool eight years into four seasons instead of mapping,
    say, just August 2024 across all states? (Look back at how sparse
    the single-year, single-month cells would have been.)

------------------------------------------------------------------------

# Part 7 — Linking the environmental signal to the clinical outcome

Now the environmental pharmacoepi question: is a state's environmental
surveillance signal for *Salmonella* associated with its clinical MDR
*E. coli* burden? We'll use the full 2018-2025 total this time as we just saw in 
parts 5-6 just  a single year can be noisy.

```{r total-env-exposure}
env_total <- beam_panel %>%
  group_by(stusps) %>%
  summarize(env_isolates_total = sum(isolates), .groups = "drop")

sum(env_total$env_isolates_total == 0)   # states with zero isolates, 2018-2025 combined
```

```{r scatter}
plot_df <- map_data %>%
  st_drop_geometry() %>%
  left_join(env_total, by = "stusps")

ggplot(plot_df, aes(x = env_isolates_total, y = percent_resistant)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "firebrick") +
  labs(x = "Environmental Salmonella isolates, 2018-2025 total",
       y = "% MDR E. coli among HAIs, 2024",
       title = "Multi-year environmental surveillance signal vs. clinical resistance burden") +
  theme_minimal()

cor.test(plot_df$env_isolates_total, plot_df$percent_resistant, method = "spearman")
```

**Discussion of the exercise. Work through it as a group.**

1.  Is there a correlation? Is it what you expected? Compare the number
    of zero-isolate states here (pooling eight years) to the single-year
    2024 map from Part 4 — pooling helped, but did it fully solve the
    surveillance-effort problem, or just shrink it?
2.  This is still **surveillance/detection bias**: an "exposure" built
    from surveillance counts is measuring exposure *and* testing effort
    at the same time, and you generally can't fully separate the two
    from counts alone.
3.  Both variables here are aggregated to the *state* level, an
    administrative boundary chosen for governance, not for epidemiology.
    Two pieces of jargon worth knowing cold for this field:
    - **Ecological fallacy**: even if the correlation above is real, it
      tells you nothing about *individual* patients — you cannot
      conclude that a specific person's resistant infection was caused
      by environmental exposure just because their state shows an
      association.
    - **Modifiable Areal Unit Problem (MAUP)**: had we aggregated to
      counties, HHS regions, or watersheds instead of states, both the
      magnitude and even the direction of this relationship could
      change, simply from redrawing the boundaries — with no change to
      the underlying reality.
4.  If you were designing a real study on this question, what exposure
    variable would you actually want, and at what spatial *and temporal*
    resolution?


# Wrap-up: taking this back to your own research

A few closing questions to carry out of the room:

- Everything today used pre-aggregated, area-level administrative data.
  What would change in this workflow if you instead had **point-level**
  data — patient addresses, well locations, facility coordinates? (Think
  about buffers, distance-to-nearest-source, and kernel density instead
  of choropleths.)
- Where in your own work do you currently use a "zone" (ZIP code,
  county, census tract, state) mostly because it's the unit your data
  happens to come in, rather than because it's the right scale for the
  exposure mechanism you care about?
- Surveillance effort bias showed up clearly in Parts 5-7. Where might it
  be hiding, less visibly, in a dataset you already use?

## Where to go next: spatial clustering

Today's exercise stopped at "here's the pattern, and here's why you
should be cautious about it." A natural next question — is a pattern you
can *see* on a map actually clustered more than chance would produce? —
is answered with spatial autocorrelation statistics (global and local
Moran's I, using the `spdep` package). That's a self-contained follow-on
exercise, deliberately not covered today so we could spend the time on
these more foundational skills. Ask your instructors for it if you want
to keep going on your own after the course.

## Further resources

- Pebesma & Bivand, *Spatial Data Science* (free online) — the standard
  modern reference for `sf`
- `tigris` documentation for other Census geographies (counties, ZCTAs,
  tracts) — the natural next step beyond state-level analysis
- `spdep` package vignettes, if/when you pick up the spatial clustering
  follow-on exercise above

